---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r LoadData}
activitydata<-read.csv("activity.csv")
```

Load the data by reading in the .csv file.

```{r StrData}
str(activitydata)
```

If we look at the structure of the data, we see that the Steps Variable is Integer, the Date Variable is a Factor, and the Interval Variable is also Integer. For now, we will keep things formatted as they are.

## What is mean total number of steps taken per day?

```{r SumStepsByDay}
TotalStepsDaily<-tapply(activitydata$steps,activitydata$date,sum,na.rm=TRUE)
hist(TotalStepsDaily,main="Histogram of Total Daily Steps", xlab="Total Steps Daily")
```

If we take the sum of Steps by Date (ignoring the NA values), and plot the Histogram of those Totals, we can see the frequency distribution of Total Steps Daily.

```{r MeanTotalStepsDaily}
mean(TotalStepsDaily)
```
```{r MedianTotalStepsDaily}
median(TotalStepsDaily)
```

Taking the Mean of the Total Steps Daily, we get `MeanTotalStepsDaily`, and taking the Median we get `MedianTotalStepsDaily`.

## What is the average daily activity pattern?

```{r AverageStepsInterval}
AverageStepsInterval<-tapply(activitydata$steps,activitydata$interval,mean,na.rm=TRUE)
plot(x=names(AverageStepsInterval),y=AverageStepsInterval,type="l",main="Average Steps taken by Interval",xlab="Interval",ylab="Average Steps Taken")
```

Next, we take the Mean of steps by Interval (again ignoring the NA values), and plot the Average Steps taken by Interval as a time series plot.

But which Interval has the highest Average Number of Steps taken?

```{r WhichIntervalMax}
names(which.max(AverageStepsInterval))
```

We can see that the Interval with the highest Average Number of Steps is Interval `WhichIntervalMax`.

## Imputing missing values

Now we want to see if imputing values into the NA observations will give us the same results.

First -- how many NAs are present in our Steps Variable?

```{r HowManyNAs}
sum(is.na(activitydata$steps))
```

There are `HowManyNAs` NA entries in the Steps Variable.

Let's fill in those NA entries with the average Steps for the Interval of each of those observations. We'll do this with a duplicate of our ActivityData data frame and then plot a histogram of the results.

```{r ImputeData}
activitydataImputed<-activitydata
activitydataImputed$steps<-with(activitydata,ave(steps,interval,FUN=function(x) replace(x,is.na(x),mean(x,na.rm=TRUE))))
TotalStepsDailyImputed<-tapply(activitydataImputed$steps,activitydataImputed$date,sum,na.rm=TRUE)
hist(TotalStepsDailyImputed,main="Histogram of Total Daily Steps with Imputed Values", xlab="Total Steps Daily")
```

Looking at the Frequency Distribution of our Imputed data, we can see that it looks a little different from our first Histogram of the Total Daily Steps where we ignored the NAs.

So how do our new Mean and Median compare to our first values?

```{r MeanImputed}
mean(TotalStepsDailyImputed)
```
```{r MedianImputed}
median(TotalStepsDailyImputed)
```

Taking the Mean of our Total Steps Daily including the Imputed values, we get `MeanImputed`, and taking the Median we get `MedianImputed`.

Compare those to our original results -- ignoring the NAs instead of imputing them, we had a Mean of `MeanTotalStepsDaily`, and a Median of `MedianTotalStepsDaily`.

## Are there differences in activity patterns between weekdays and weekends?

To see if there are differences between Weekdays and Weekends, first we need to create a Factor Variable that labels each observation as "Weekday" or "Weekend", based on the date. Since we did not pre-proces our Date Variable, we will need to make sure we convert that Variable to a Date when we do this.

```{r CreateFactor}
Weekdays<-weekdays(as.Date(as.character(activitydataImputed$date)))
WeekdaysFactor<-factor(Weekdays,levels=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))
activitydataImputed$DayOfWeek<-factor(unclass(WeekdaysFactor)<=5,levels=c(FALSE,TRUE),labels=c("Weekend","Weekday"))
```

Now that we've labeled each observation according to whether they occurred on a weekday or a weekend, we can take a look at our Average Steps by Interval between the two.

```{r PlotByDayOfWeek}
AverageStepsIntervalDayOfWeek<-aggregate(steps~interval+DayOfWeek,data=activitydataImputed,mean)
xyplot(steps~interval|DayOfWeek,data=AverageStepsIntervalDayOfWeek,layout=c(1,2),main="Average Steps Taken by Interval",xlab="Interval",ylab="Average Steps Taken",type="l")
```

It looks like there is a difference between the two after all - especially around the peak Interval we saw before, Interval `WhichIntervalMax`.